---
title: "Basics: making and using bathymetric maps with marmap2"
output: rmarkdown::html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "",
  fig.width = 9,
  fig.height = 4.8,
  fig.retina = 2,
  dpi = 160,
  message = FALSE,
  warning = FALSE
)

options(
  cli.num_colors = 1,
  crayon.enabled = FALSE,
  pillar.bold = FALSE,
  pillar.subtle = FALSE
)
```


```{r local-data, include = FALSE}
load_example <- function(file) {
  tibble::as_tibble(readRDS(system.file("extdata", file, package = "marmap2")))
}

alboran <- load_example("alboran_tbl.rds")
aleutians <- load_example("aleutians_tbl.rds")
irregular <- load_example("irregular_tbl.rds")
medit <- load_example("mediteranean_tbl.rds")
```

# Introduction

`marmap2` provides tools for importing, transforming and plotting bathymetric
and hypsometric data in R. It keeps the main workflow close to modern R habits:
data are returned as tidy long tables by default, can be manipulated with pipes,
and can be plotted with ggplot2. Compatibility with the historical `bathy`
matrix format is still available when older marmap code needs it.

In this vignette, we will download bathymetric grids, inspect and simplify them,
build maps one layer at a time, compare bathymetric colour scales, work around
the Pacific antimeridian, project a Mediterranean grid, regularise irregularly
spaced soundings, and convert bathymetric data to other spatial formats.

# A Quick Tutorial

The examples below use several regions with contrasted topography and
bathymetry: the Alboran Sea, the Mediterranean Sea, the Aleutian arc and a set
of irregularly spaced Hudson Bay soundings. The download commands are shown
exactly as they would be used in an interactive R session.

```{r packages}
library(marmap2)
library(ggplot2)
```

## Getting Data Into R

The simplest way to get data into `marmap2` is to request a geographic window from either [GEBCO](https://www.gebco.net) or [NOAA](https://www.ngdc.noaa.gov/mgg/bathymetry/bathymetry.html). GEBCO currently returns its native 15 arc-second grid. This is a good source when a detailed local map is needed.

```{r download-gebco, eval = FALSE}
alboran <- get_gebco(
  lon = c(-6, 0),
  lat = c(35, 37)
)
```

For detailed maps covering limited areas, NOAA data can be loaded at a resolution
comparable to GEBCO. For maps covering large geographic extents, the `resolution` argument of `get_noaa()` can be used to request a coarser grid and avoid (down)loading unnecessarily large datasets. The `resolution` argument is expressed in arc-minutes. The highest resolutions are obtained with `resolution = 0.25` (0.25 arc-minutes, equivalent to 15 arc-seconds).

```{r download-noaa, eval = FALSE}
medit <- get_noaa(
  lon = c(-6, 36.5),
  lat = c(30, 46),
  resolution = 4
)
```

Both `get_gebco()` and `get_noaa()` provide a `keep` argument to store a local copy of the downloaded data in the working directory. This helps avoid repeated requests to the NOAA and GEBCO servers when the same dataset is used several times.

The workflow remains unchanged: when data are requested with either function, the working directory is checked first. If the corresponding dataset is already available locally, this copy is loaded directly; otherwise, the data are downloaded from the online source. The directory used to save and retrieve local files can be specified with the `path` argument of both functions.

`summarise_bathy()` can be used to print useful information about the imported data.


```{r inspect-data}
head(alboran)

summarise_bathy(alboran)

summarise_bathy(medit)
```

If you have already downloaded a high-resolution grid but only need a lighter object for mapping or computation, you can reduce the resolution locally with `reduce_bathy_resolution()`. This is preferable to sending repeated queries for the same area to a remote server. As in get_noaa(), the `resolution` argument is expressed in arc-minutes, and the `method` argument specifies how values in the coarser grid are derived from the finer grid: nearest-cell value (default), mean value, or median value.

```{r reduce-resolution}
alboran_5 <- reduce_bathy_resolution(
  alboran,
  resolution = 5
)

summarise_bathy(alboran_5)
```

Both `get_noaa()` and `get_gebco()` return a tibble by default. `class = "bathy"` can be used to return a legacy bathy matrix. A tibble can also be converted to a legacy bathy matrix with `tbl_to_bathy()` after import.

```{r legacy-bathy}
alboran_bathy <- tbl_to_bathy(alboran)
class(alboran_bathy)
```

## Plotting Bathymetric Data

A first map can be produced in one line. `quickplot_bathy()` is meant for data
inspection just after downloading or transforming a grid.

```{r quickplot, fig.width = 9, fig.height = 4.2}
quickplot_bathy(alboran)
quickplot_bathy(alboran_5)
quickplot_bathy(medit)
```

The same maps can be built layer by layer. `geom_bathy()` draws the bathymetric
raster, `geom_coastline()` draws the 0 m contour, and `geom_contour()` can add
isobaths and elevation contours. The manual appraoch allows for finer control over the appearance of each layer and the overall map. The example below shows a detailed map of the Alboran Sea, with bathymetry, coastline, isobaths every hundred meters and elevation contours every 500 meters.

```{r detailed-alboran-map, fig.width = 9, fig.height = 4.4}
alboran |>
  ggplot() +
  geom_bathy(expand = FALSE) +
  geom_coastline(linewidth = 0.45) +
  geom_contour(
    aes(lon, lat, z = depth),
    breaks = seq(0, -3000, by = -100),
    colour = "grey25",
    linewidth = 0.18
  ) +
  geom_contour(
    aes(lon, lat, z = depth),
    breaks = seq(0, 3000, by = 500),
    colour = "grey25",
    linewidth = 0.18
  ) +
  scale_fill_bathy(
    palette_ocean = "ocean_blues",
    palette_land = "land_hcl"
  ) +
  theme_bw()
```

The palette system is deliberately simple: provide one palette for the ocean and
one for land. A palette can be one of the built-in names (try
`bathy_palettes()` for the full list), a vector of colours, a palette function,
or a single colour. This makes it possible to use the palettes shipped with
`marmap2`, palettes from other R packages, base R HCL palettes, or hand-made
gradients. The only important rule is to choose ocean and land palettes that
work together visually and remain ordered from deep to shallow water, and from
low to high elevation.

Here is the Alboran Sea with two palettes shipped with `marmap2`.

```{r palette-built-in, fig.width = 9, fig.height = 4.2}
alboran |>
  ggplot() +
  geom_bathy(expand = FALSE) +
  geom_coastline(linewidth = 0.4) +
  scale_fill_bathy(
    palette_ocean = "ocean_mako",
    palette_land = "land_earth",
    mode = "rescale"
  ) +
  labs(title = "Built-in marmap2 palettes") +
  theme_bw()
```

The same function can use colour-generating functions. The next example uses
base R HCL palettes through `grDevices::hcl.colors()`.

```{r palette-hcl, fig.width = 9, fig.height = 4.2}
alboran |>
  ggplot() +
  geom_bathy(expand = FALSE) +
  geom_coastline(linewidth = 0.4) +
  scale_fill_bathy(
    palette_ocean = function(n) grDevices::hcl.colors(n, "Blues 3"),
    palette_land = function(n) grDevices::hcl.colors(n, "Terrain 2"),
    mode = "rescale"
  ) +
  labs(title = "Base R HCL palettes") +
  theme_bw()
```

The same approach can mimic well-known perceptual palettes without adding
runtime dependencies. Here, two HCL palettes are turned into explicit colour
vectors before being passed to `scale_fill_bathy()`.

```{r palette-external, fig.width = 9, fig.height = 4.2}
ocean_viridis_like <- grDevices::hcl.colors(9, "Viridis")
land_ylorbr_like <- grDevices::hcl.colors(9, "YlOrBr")

alboran |>
  ggplot() +
  geom_bathy(expand = FALSE) +
  geom_coastline(linewidth = 0.4) +
  scale_fill_bathy(
    palette_ocean = ocean_viridis_like,
    palette_land = land_ylorbr_like,
    mode = "rescale"
  ) +
  labs(title = "Dependency-free HCL colour vectors") +
  theme_bw()
```

Finally, a palette can be created manually from colour vectors. This is useful
when preparing several figures that need a specific visual identity.

```{r palette-manual, fig.width = 9, fig.height = 4.2}
alboran |>
  ggplot() +
  geom_bathy(expand = FALSE) +
  geom_coastline(linewidth = 0.4) +
  scale_fill_bathy(
    palette_ocean = c("#081D58", "#225EA8", "#41B6C4", "#C7E9B4", "#F7FCF0"),
    palette_land = c("#F6E8C3", "#D8B365", "#8C510A", "#4D2D17"),
    mode = "rescale"
  ) +
  labs(title = "Manual ocean and land gradients") +
  theme_bw()
```

The `mode` argument controls how colours are assigned to depth and elevation
values. With `mode = "truncate"`, colours remain anchored to fixed reference
depths and elevations: the same depth keeps the same colour across different
maps. The reference can be supplied manually with a numeric vector, or more
simply with a bathymetric object through the `reference_limits` argument. With
`mode = "rescale"`, the complete palette is stretched over the range of the
plotted data, which maximises local contrast but makes colours less directly
comparable between maps.

The Mediterranean map below defines the regional reference. Because
`reference_limits = medit`, the complete palette is used across the depth and
elevation range of the Mediterranean dataset.

```{r palette-mediterranean-truncate, fig.width = 9, fig.height = 4.2}
medit |>
  ggplot() +
  geom_bathy(expand = FALSE) +
  geom_coastline(linewidth = 0.35) +
  geom_contour(
    aes(lon, lat, z = depth),
    breaks = seq(0, -5000, by = -500),
    colour = "grey20",
    linewidth = 0.14
  ) +
  scale_fill_bathy(
    palette_ocean = "ocean_blues",
    palette_land = "land_earth",
    mode = "truncate",
    reference_limits = medit
  ) +
  labs(title = "Mediterranean Sea, regional reference") +
  theme_bw()
```

The two Alboran maps use the same palettes but different modes. In truncate
mode, the Alboran colours match the Mediterranean colours for the same depths.
In rescale mode, the full palette is used within the local depth range of the
Alboran Sea.

```{r palette-alboran-truncate, fig.width = 9, fig.height = 4.2}
alboran |>
  ggplot() +
  geom_bathy(expand = FALSE) +
  geom_coastline(linewidth = 0.4) +
  geom_contour(
    aes(lon, lat, z = depth),
    breaks = seq(0, -3000, by = -250),
    colour = "grey20",
    linewidth = 0.16
  ) +
  scale_fill_bathy(
    palette_ocean = "ocean_blues",
    palette_land = "land_earth",
    mode = "truncate",
    reference_limits = medit
  ) +
  labs(title = "Alboran Sea, truncate mode, Mediterranean reference") +
  theme_bw()
```

```{r palette-alboran-rescale, fig.width = 9, fig.height = 4.2}
alboran |>
  ggplot() +
  geom_bathy(expand = FALSE) +
  geom_coastline(linewidth = 0.4) +
  geom_contour(
    aes(lon, lat, z = depth),
    breaks = seq(0, -3000, by = -250),
    colour = "grey20",
    linewidth = 0.16
  ) +
  scale_fill_bathy(
    palette_ocean = "ocean_blues",
    palette_land = "land_earth",
    mode = "rescale"
  ) +
  labs(title = "Alboran Sea, rescale mode") +
  theme_bw()
```

For a contour-only map, you can use `scale_colour_bathy()` with the levels
computed by `geom_contour()` to color code isobaths or elevation contours. In
the example below, `geom_bathy()` is still used to draw the geographic
background and land mask, but the ocean fill is made transparent. The depth
information is carried only by the contour lines.

```{r contour-only, fig.width = 9, fig.height = 4.2}
alboran |>
  ggplot(aes(lon, lat, z = depth)) +
  geom_bathy(expand = FALSE, show.legend = FALSE) +
  geom_contour(
    aes(colour = after_stat(level)),
    breaks = seq(-2500, 0, by = 500),
    linewidth = 0.5
  ) +
  geom_coastline() +
  scale_fill_bathy(
    palette_ocean = "transparent",
    palette_land = "grey80",
    mode = "rescale",
    name = "depth"
  ) +
  scale_colour_bathy(
    palette_ocean = "ocean_blues",
    palette_land = NULL,
    mode = "rescale",
    name = "depth"
  ) +
  theme_bw()
```

Here, `scale_fill_bathy()` controls the background only: ocean cells are
transparent and land cells are grey. `scale_colour_bathy()` controls the
isobaths. Setting `palette_land = NULL` tells the colour scale that only ocean
values should be mapped, so shallow negative depths are not treated as land
colours. `show.legend = FALSE` in `geom_bathy()` avoids displaying a second
legend for the transparent background layer.

## Preparing Maps In The Pacific Antimeridian Region

A region that crosses the antimeridian is best requested explicitly. For
example, the Aleutian arc spans longitudes east and west of 180 degrees.

```{r antimeridian-download, eval = FALSE}
aleutians <- get_noaa(
  lon = c(165, -145),
  lat = c(50, 65),
  resolution = 5,
  antimeridian = TRUE
)
```

When the data use continuous longitudes around 180 degrees, set
`antimeridian = TRUE` in `geom_bathy()`. The x axis then remains readable, with
labels expressed as east and west longitudes.

```{r antimeridian-plot, fig.width = 9, fig.height = 3.9}
aleutians |>
  ggplot() +
  geom_bathy(
    aes(lon, lat, fill = depth),
    antimeridian = TRUE,
    expand = FALSE,
    x_breaks = seq(165, 215, by = 10)
  ) +
  geom_coastline(linewidth = 0.4) +
  geom_contour(
    aes(lon, lat, z = depth),
    breaks = seq(0, -7000, by = -500),
    colour = "grey20",
    linewidth = 0.15
  ) +
  scale_fill_bathy(
    palette_ocean = "ocean_blues",
    palette_land = "land_earth"
  ) +
  theme_bw()
```

## Projections

Geographic longitude/latitude grids are convenient for downloading data, but a
projected coordinate reference system is often preferable for regional maps.
Here, a Mediterranean grid is downloaded from NOAA, projected to ETRS89 / LAEA
Europe (EPSG:3035), and plotted in one pipe.

```{r projection-download, eval = FALSE}
medit <- get_noaa(
  lon = c(-6, 36.5),
  lat = c(30, 46),
  resolution = 4
)
```

```{r projection-map, fig.width = 9, fig.height = 4.6}
medit |>
  project_bathy(
    crs_to = 3035,
  ) |>
  ggplot() +
  geom_bathy(expand = FALSE) +
  geom_coastline(linewidth = 0.35) +
  scale_fill_bathy(
    palette_ocean = "ocean_blues",
    palette_land = "land_earth"
  ) +
  labs(x = "Easting", y = "Northing") +
  theme_bw()
```

The visible grid is no longer a simple longitude/latitude rectangle: the
projection and raster resampling have changed the shape of the mapped region.

## Irregularly Spaced Data

Bathymetric data are sometimes available as irregularly spaced soundings rather
than as a regular grid. `griddify()` bins these points into a regular grid and,
optionally, fills empty cells by inverse distance weighting.

```{r griddify, fig.width = 6.2, fig.height = 5.4}
set.seed(1)
irregular_sample <- irregular[sample(nrow(irregular), 1600), ]

hudson <- griddify(
  irregular_sample,
  nlon = 90,
  nlat = 90,
  interpolate = TRUE
)

hudson |>
  ggplot() +
  geom_bathy(expand = FALSE) +
  geom_point(
    data = irregular_sample,
    aes(lon, lat),
    inherit.aes = FALSE,
    size = 0.25,
    alpha = 0.35
  ) +
  scale_fill_bathy(
    palette_ocean = "ocean_mako",
    palette_land = NULL,
    mode = "rescale",
    limits = c(min(hudson$depth, na.rm = TRUE), 0)
  ) +
  theme_bw()
```

The points show the original sampling pattern. The background grid is the
regularised bathymetry returned by `griddify()` and can be passed to all the
same plotting and conversion functions as imported data.

## Other Spatial Formats

The default tibble format is convenient for ggplot2 and tidy workflows. Other
formats are available when needed.

```{r conversions}
alboran_bathy <- tbl_to_bathy(alboran)
alboran_tbl <- bathy_to_tbl(alboran_bathy)
alboran_xyz <- as_xyz(alboran_tbl)
alboran_sf <- as_sf(alboran_tbl)
alboran_raster <- as_spatraster(alboran_tbl)

class(alboran_bathy)
class(alboran_tbl)
class(alboran_xyz)
class(alboran_sf)
class(alboran_raster)
```

Use `tbl_to_bathy()` when older marmap code expects a `bathy` matrix. Use
`as_xyz()` for external software that expects a plain three-column xyz table.
Use `as_sf()` and `as_spatraster()` to work with the wider R spatial ecosystem.
